---
title: "Choices"
---
```{r, setup, include=FALSE}
source("setup.R")
```
## The `if` statement
Choices enable you to run different code based on conditions / input.
The most basic form of choices is the `if` statement.

::: {.panel-tabset}

## Python
Note, that all
```{python}
z = 4
if z % 2 == 0:
    print("z is even")
else:
    print("z is uneven")
```

## R
in one line
```{r}
z = 4
if (z %% 2 == 0) print("z is even") else print("z is odd")
```

in several lines
```{r}
z = 5
if (z %% 2 == 0) {
  print("z is even")
} else {
  print("z is odd")
}
```

## SQL
Note, that SQLite does not support the "normal" `if` statement. However, the same can be achieved with the `case when` statement.
```{sql}
#| connection: con
-- SQLite flavor
select
  price
  ,  case when price >= 500 then 'high-price' else 'low-price' end as price_categ
from
  (select 750 as price)
limit 5;
```
## Bash
```{bash}
#!/bin/bash

z=4

if [ $((z % 2)) -eq 0 ]; then
    echo "z is even"
else
    echo "z is odd"
fi
```
:::

## The `elif` statement
For more complex choices, the `elif` or `else if` statement can be used.

:::{.callout-caution}
Keep in mind, that only the line of code in the **first** `ture` condition will run. The order of conditions matters!
:::

::: {.panel-tabset}

## Python
```{python}
def get_price_categ(price):
  categ = str()
  if price > 1000:
    categ = "high"
  elif price > 700:
    categ = "middle"
  elif price > 300:
    categ = "low"
  elif price > 0:
    categ = "super cheap"
  elif price == 0:
    categ = "free"
  else:
    categ = "there must be an error"
  return categ

get_price_categ(750)  
```

## R
The `if` in R works specifically or scalars.

```{r}
get_price_categ <- function(price) {
  categ <- character()
  if (price > 1000) {
    categ <- "high"
  } else if (price > 700) {
    categ <- "middle"
  } else if (price > 300) {
    categ <- "low"
  } else if (price > 0) {
    categ <- "super cheap"
  } else if (price == 0) {
    categ <- "free"
  } else {
    categ <- "there must be an error"
  }
  return(categ)
}

get_price_categ(17)
```

## SQL
Note, that many SQL flavors do not have a classic `elif`, but the `case when` can be used as an `elif` like the following:
```{sql}
#| connection: con
-- SQLite flavor
select 
  price,
  case 
    when price > 1000 then 'high'
    when price > 700 then 'middle'
    when price > 300 then 'low'
    when price > 0 then 'super cheap'
    when price = 0 then 'free'
    else 'there must be an error'
  end as price_category
from
  (select 750 as price) as subquery;
```
:::

## The `if` / `elif` / `case when` statement for vectors / columns
A classic task in data cleaning is changing the values of a column in a df to a different value, based on conditions.

::: {.panel-tabset}
## Python
```{python}
#| eval: false
import pandas as pd
df = orders[['quantity']].copy()
df['order_size'] = df['quantity'].apply(lambda x: 'large' if x >= 3 else 'small')
df = df.iloc[0:5]

print(df)

```

## R
The `ifelse` in R works specifically for vectors.
```{r}
df <- orders %>% 
  select(quantity) %>% 
  mutate(order_size = ifelse(quantity > 3, "large", "small")) %>% 
  slice(1:5)
df
```
For more conditions, the `dplyr::case_when` feels more natural and improves readability, especially when having several conditions
```{r}
df <- orders %>%
  select(quantity) %>% 
  mutate(
    quantity_categ = case_when(
      quantity == 1 ~ "small",
      quantity %in% c(2, 3) ~ "middle",
      quantity > 3 ~ "large",
      TRUE ~ "there must be sth wrong" # this line is like an "else"
    )
  ) %>% 
  slice(1:5)
df
```

## SQL
```{sql}
#| connection: con
-- SQLite flavor
select quantity,
    case
        when quantity = 1 then 'small'
        when quantity IN (2, 3) then 'middle'
        when quantity > 3 then 'large'
        else 'there must be sth wrong'
    end as quantity_categ
from orders
limit 5;
```
:::

## The `switch` statement
The switch statement is especially useful for getting values from a dictionary-like construct with key and value.

::: {.panel-tabset}
## R
```{r}
get_height <- function(x) {
  switch(x,
    "eiffel tower" = 330,
    "Burj Khalifa" = 828,
    "Shanghai Tower" = 632,
    stop("not found") # always include a "not found" for debugging purposes
  )
}
get_height("Shanghai Tower")
```
:::