---
title: "Loops"
---
```{r, setup, include=FALSE}
source("setup.R")
```
## The `for` loop
`for` loops are probably the most used form of loops. A for loop iterates over iterable objects such as lists, vectors, sets and other data structures. In contrast to the `while` loop, the `for` loop is less prone to ending up in infinite loop and is therefore preferable.
:::{{.callout-note}}
Note, that for loops can often be replaced with the more sophisticated apply, sapply or lapply.
:::

:::{.panel-tabset}

## Python
Iterate over the items in a list:
```{python}
x = ["hello", "my", "name", "is", "eugen", "what's", "your", "name?"]
for i in x:
    print(i)
```
Get the index of the element using enumerate():
```{python}
x = ["hello", "my", "name", "is", "eugen", "what's", "your", "name?"]
for i, value in enumerate(x):
    print(f"{i} {value}")
```

:::{.callout-caution}
Keep in mind, that iterables start with the index `0` in python
:::

Jump to next iteration using `continue`, exit the loop using `break`:
```{python}
x = ["hello", "my", "name", "is", "eugen", "what's", "your", "name?"]
for i in x:
    if i in ["my", "name", "is"]:
        continue
    else:
        print(i)
        if i == "eugen":
            break
```

## R
Iterate over the items in a vector:
```{r}
x = c("hello", "my", "name", "is", "eugen", "what's", "your", "name?")
for (i in x) {
  print(i)
}
```
Get the index of the element using seq_along():
```{r}
x = c("hello", "my", "name", "is", "eugen", "what's", "your", "name?")
for (i in seq_along(x)) { # always use seq_along instead of 1:length(x) or 1:nrow(df)
  print(paste(i,x[i]))
}
```

:::{.callout-caution}
Keep in mind, that iterables start with the index `1` in R.
:::

Jump to next iteration using `next`, exit the loop using `break`:
```{r}
x = c("hello", "my", "name", "is", "eugen", "what's", "your", "name?")
for (i in x) {
  if (i %in% c("my", "name", "is")) {
    next
  } else {
    print(i)
    if (i == "eugen") break
  }
}
```

<!-- ## Bash -->
<!-- ```{bash} -->
<!-- #!/bin/bash -->

<!-- x=("hello" "my" "name" "is" "eugen" "what's" "your" "name?") -->

<!-- for i in "${x[@]}" -->
<!-- do -->
<!--     echo "$i" -->
<!-- done -->
<!-- ``` -->

:::
